---
title: "“Introduction to ggplot2""
author: "Sarah Supp"
date: "5/16/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


**There is a presentation which goes with this handout, use it to add notes if you need to.**

Before we start, remember we need to do a couple of things...


#### 1. Tell R where your data are
See the introduction handout if you can't remember how.

Check it worked using:
```{r}
list.files()
```
or look at the Files tab

#### 2. Load the packages needed
If you downloaded the entire "tidyverse" then you should already have the ggplot2 package. If not, then you will need to get it using "install.packages"
```{r}
library(tidyverse)
library(ggplot2)
```

#### 3. Import the data
First we are going to work with the dataset `MOMv3.3.txt`, which contains the body mass of all mammals on Earth (MOM = Masses Of Mammals). You can find the original source by Smith et al. 2003, [here](http://esapubs.org/archive/ecol/E084/094/)
```{r}
MOM <- read_tsv("MOMv3.3.txt", col_names=F)

#Let's name the columns so they are easier to work with
names(MOM) <- c("continent", "status", "order", "family", "genus", "species", "log_mass_g", "avg_mass_g", "reference")
```

#### 4. Look at the data
```{r}
glimpse(MOM)
```

### What plots do I need?
Before we start to make plots, let's think about what we'd like to do more carefully.

1. What kind of data do we have? e.g. quantitative? qualitative?
2. What is the main question we'd like to answer?
3. What would be interesting to show about our data?

We can spend all day making graphs, but there are a few rules of thumb we can use to avoid wasting effort, and to avoid going down a rabbit hole that distracts us from our main purpose.

1. What's your point? 
Are you visualizing data for exploratory purposes (to understand it better) or for explanatory purposes (to make a specific point to someone else)?
The goal is not always to display *all* of the data as neatly and completely as possible!

2. Choose the right plot
Think about what each kind of plot can tell you and how your data type constrains your options.
Would quantitative or qualitative data be best shown on this plot?
Would the plot be best to display discrete data (categories, integers) or continuous data?
Does this plot show all the data, or does it break it into distinct groups?

* histograms can show the *distribution* of 1 quantitative variable
* bar charts or boxplots can *compare* quantities in 2 or more different categories
* scatterplots show the *relationship* of two items
* line charts can track *changes through time*, or show the relationship of 2 or more variables
* pie charts compare parts of a whole
* facets can *divide information into distinct groups*
* layers can add more information as *color, shape, size* etc.
* advanced plotting commands may include: heatmaps, layered or stacked bar graphs, maps, animated plots, 3D plots, and more!

3. Less is more
Don't overcomplicate and try to be as straightforward as possible. Especially for explanatory plots, you want to try to focus your reader's attention.

4. Use color intentionally
Use color to emphasize certain parts of a graph, or to push other parts to the background. For example, color could differentiate groups in a graph, or help your key results that are the most exciting to stand out!

5. Add a title and good axis labels to help read and understand your graph

6. Experiment, get feedback, and iterate - can someone else understand it?

There are **many** other options and different opinions on making graphics. Let's start with some of the most common ones, and make some plots!


