---
title: "intro-ggplot2"
author: "Sarah Supp"
date: "5/31/2018"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(tidyverse)
library(naniar)

# Masses of Mammals
MOM <- read_tsv("MOMv3.3.txt", col_names=F)
names(MOM) <- c("continent", "status", "order", "family", "genus", "species", "log_mass_g", "avg_mass_g", "reference")
MOM <- MOM %>%
  replace_with_na(replace = list(avg_mass_g = -999, log_mass_g = -999.00))

# Climate data
clim <- read_csv("climate_anomalies_1880-2016.csv")

```

## Challenge 1
Make a boxplot like the one we just did, but try adding layers for axis labels, color, or to compare different categories (`facet_wrap` or change the x-variable). 
```{r, warning=FALSE}
ggplot(MOM, aes(x = continent, y = log_mass_g)) + 
  geom_boxplot()
```


## Challenge 2
Can you create the plot shown below? 
(Hint: if you have imported ggplot2, you already have the "iris" dataset)

```{r}
ggplot(iris, aes(Sepal.Length, Sepal.Width, color = Species)) +
  geom_point(aes(shape = Species), size = 3)
```


## Challenge 3
Create the plot below using 1000 points from the `diamonds` dataset.
(Hint: Use `sample_n()` to make a smaller dataset). For an added challenge, make the points semi-transparent.

```{r}
d2 <- diamonds %>% sample_n(1000)
ggplot(d2, aes(carat, price)) + geom_point(aes(col=color))
```


## Challenge 4
You can add a trendline to a plot using stat_smooth(). Try making the plot below with the `iris` dataset, with a trendline and color added.
```{r, warning=FALSE,message=FALSE}
ggplot(iris, aes(Sepal.Width, Petal.Length, group=Species, col=Species)) + stat_smooth(method="lm")
```

## Challenge 5

Try making the plot below, by modifying the climate plot we just made:

```{r, message=FALSE, warning=FALSE}
ggplot(clim, aes(Date, Mean, group=Source, col=Source)) + stat_smooth() + geom_hline(yintercept=0, linetype="dashed")
```

