---
title: "Dinosaur Super Challenge"
author: "Sarah Supp"
date: "6/7/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Sarah Supp
NHM Course, May 2018

## Import the libraries and data needed
```{r, warning=FALSE, echo=FALSE}
library(tidyverse)

dinos <- read_csv("example_data/dino_collection.csv") 

```

***

## Write functions to help calculate mass from length and deal with different taxa

estimate dinosaur mass (kg) from length (m)
```{r}
est_mass_from_length <- function(length, a, b){
  # dinosaur mass, using a and b; Seebacher 2001
  mass <- a * length ** b
  return(mass)
}
```

convert mass (kg) to weight (lbs)
```{r}
kg2lbs <- function(kg) {
  # converts kilograms to pounds
  pounds <- kg * 2.205
  return(pounds)
}
```

Identify the suborder from the dataset and add columns for a and b parameters
```{r}

find_suborder_params <- function(data){
  #identify suborder and correct a, b parameters
  # input: dataset with columns for suborder and length
  # output: dataset with columns for a and b parameters
  data_with_params <- data %>%
    mutate(a = case_when(suborder == "therapoda" ~ 0.73,
                         suborder == "sauropoda" ~ 214.44,
                         suborder == "ankylosauridae" ~ 16.54,
                         suborder == "ceratopsia" ~ 12.58,
                         suborder == "stegosauria" ~ 10.95,
                         suborder == "ornithopoda" ~ 11.81,
                         suborder == "prosauropoda" ~ 12.32,
                         TRUE ~ NA_real_)) %>%     
            # the above line is for an instance where "case_when" doesn't match anything
    mutate(b = case_when(suborder == "therapoda" ~ 3.63,
                         suborder == "sauropoda" ~ 1.46,
                         suborder == "ankylosauridae" ~ 2.51,
                         suborder == "ceratopsia" ~ 2.90,
                         suborder == "stegosauria" ~ 2.64,
                         suborder == "ornithopoda" ~ 2.66,
                         suborder == "prosauropoda" ~ 2.40,
                         TRUE ~ NA_real_))
  return(data_with_params)
}
```

***

## Clean and wrangle the dinosaur data
```{r}
glimpse(dinos)
summary(dinos)

dinos$suborder <- tolower(dinos$suborder)
unique(dinos$suborder)

# fix the typos; should be only 7 suborders
dinos2 <- dinos %>%
  mutate(suborder = ifelse(suborder=="tegosauria", "stegosauria", ifelse(suborder=="onithopoda", "ornithopoda", suborder)))

unique(dinos2$suborder)
```

## Calculate mass in kilograms and pounds, depending on the suborder
Add these to the existing data frame so we can plot them later
```{r}
dinos3 <- find_suborder_params(dinos2)

dino_masses<- dinos3 %>%
  mutate(mass_kg = round(a * length^b, 2),
         mass_lbs = round(kg2lbs(mass_kg),2))

```


## Plot the data
Plot the dinosaur masses by body length, and color the points by taxa. 
```{r}
ggplot(dino_masses, aes(length, mass_kg, group=suborder, col=suborder)) + 
  geom_point(aes(size=mass_lbs), alpha=0.5) +
  scale_y_log10() + 
  scale_x_log10() + 
  stat_smooth(method="lm")
```

```{r}
smallest_dino <- dino_masses %>%
  arrange(mass_kg) %>%
  slice(1)

largest_dino <- dino_masses %>%
  arrange(desc(mass_kg)) %>%
  slice(1)
```


The largest dinosaur is `r largest_dino$specimen_name` which is a `r largest_dino$mass_kg` kg `r largest_dino$suborder` and the smallest dinosaur is `r smallest_dino$specimen_name` which is a `r smallest_dino$mass_kg` kg `r smallest_dino$suborder`.

# RAWR!!
![http://placekitten.com/600/400](http://placekitten.com/600/400)


