---
title: "Basic ggmap tutorial"
output: html_document
---
  
  ```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**There is a presentation which goes with this handout, use it to add notes if you need to.**

FYI: This doesn't compile well with `Knit` due to the vagueries of Google, internet connections and R all working together. So just work with the `.Rmd`
  
Before we start, remember we need to do a couple of things...

#### 1. Tell R where your data are
See the introduction handout if you can't remember how.

Check it worked using:
```{r}
list.files()
```
or look at the Files tab

#### 2. Load the packages needed
You may need to install `ggmap` if you didn't do that already.
```{r}
library(tidyverse)
library(ggmap)
```

## ggmap magic

We first use Google maps (and the `get_map` function) to extract map data with latitude and longitude or postcodes or location names. THEN we use `ggmap` to plot the data.

**This requires an internet connection**. Note that each IP address can only download 250 a day.

```{r, message = FALSE, warning = FALSE}
mymap <- get_map(location = "Manchester, UK")
ggmap(mymap)
```

You can change the zoom...

```{r, message = FALSE, warning = FALSE}
mymap <- get_map(location = "Manchester, UK",
                 zoom = 17)
ggmap(mymap)
```

...the map type...

```{r, message = FALSE, warning = FALSE}
mymap <- get_map(location = "New York",
                 maptype = "satellite",
                 zoom = 10)
ggmap(mymap, extent = "device")
```

... the maptype and source of the map...

```{r, message = FALSE, warning = FALSE}
mymap <- get_map(location = "York",
                 source = "osm")
ggmap(mymap)
```

...use a postcode (you can find your house!)...

```{r, message = FALSE, warning = FALSE}
mymap <- get_map(location = "SW7 5BD", 
                 maptype = "satellite", 
                 zoom = 17)
ggmap(mymap)
```

...use lat long coordinates...

```{r, warning = FALSE}
mymap <- get_map(location = c(lon = 2.2945, lat = 48.8584),
                 maptype = "satellite",
                 zoom = 17)
ggmap(mymap, extent = "device")
```

...change the colour and multiple options etc. etc.

```{r, message = FALSE, warning = FALSE}
mymap <- get_map(location = c(lon = 8.5417, lat = 47.3769),
    color = "color",
    source = "google",
    maptype = "satellite",
    zoom = 15)

ggmap(mymap, extent = "device")
```

- `color` lets you set it as colour or black and white (bw)
- `source` allows you to use google, open street map etc. if you prefer
- `maptype` has lots of options, we'll explore some below.
- `zoom` tells R how zoomed in to be. 3 = continent, 10 = city, 21 = building
- See `?get_map` for more details.

`ggmap` also has options you can change:
```{r, message = FALSE, warning = FALSE}
ggmap(mymap,
    extent = "device",
    ylab = "Latitude",
    xlab = "Longitude")
```

- `extent = device` makes the whole plotting window into a map
- If you wanted to see the lat long axes and/or a legend use `extent = panel`.

Look at some of the other get_map options...

```{r, message = FALSE, warning = FALSE}
mapData1 <- get_map(location = "Berlin",
    color = "color",
    source = "google",
    maptype = "satellite",
    zoom = 13)

ggmap(mapData1,
    extent = "device",
    ylab = "Latitude",
    xlab = "Longitude")
```

### TERRAIN maps
```{r, message = FALSE, warning = FALSE}
mapData2 <- get_map(location = "Boulder",
	color = "color",
    source = "google",
    maptype = "terrain",
    zoom = 13)

ggmap(mapData2,
    extent = "device",
    ylab = "Latitude",
    xlab = "Longitude")
```

### ROAD MAP
```{r, message = FALSE, warning = FALSE}
mapData3 <- get_map(location = "Birmingham, UK",
    color = "color",
    source = "google",
    maptype = "roadmap",
    zoom = 13)

ggmap(mapData3,
    extent = "device",
    ylab = "Latitude",
    xlab = "Longitude")
```

### HYBRID maps
```{r, message = FALSE, warning = FALSE}
mapData4 <- get_map(location = "Paris",
	color = "color",
    source = "google",
    maptype = "hybrid",
    zoom = 13)

ggmap(mapData4,
    extent = "device",
    ylab = "Latitude",
    xlab = "Longitude")
```

### WATERCOLOR (useless but pretty!). Uses stamen not google.
```{r, message = FALSE, warning = FALSE}
mapData5 <- get_map(location = "London",
    color = "color",
    source = "stamen",
    maptype = "watercolor",
    zoom = 13)

ggmap(mapData5,
    extent = "device",
    ylab = "Latitude",
    xlab = "Longitude")
```

## Adding things to maps

Let's use a map of London, centred on London Bridge.

```{r, message = FALSE, warning = FALSE}
mapLondon <- get_map(location = "London Bridge",
    color = "color",
    source = "google",
    maptype = "satellite",
    zoom = 13)
```

We can add points with lat long coordinates. This uses `+ geom_point` (just like `ggplot` with layers)

This is St Paul's

```{r, message = FALSE, warning = FALSE}
ggmap(mapLondon, extent = "device") +
  geom_point(aes(x = -0.0984, y = 51.5138),
             alpha = 0.5, color = "deeppink", 
             size = 10, pch = 16)
```

We can annotate points...
```{r, message = FALSE, warning = FALSE}
ggmap(mapLondon, extent = "device")+
  geom_text(aes(x = -0.0984, y = 51.5138, label = "St Paul’s"), 
    colour = "yellow", size = 5)
```

Or do both!
```{r, message = FALSE, warning = FALSE}
ggmap(mapLondon, extent = "device") +
  geom_point(aes(x = -0.0984, y = 51.5138, 51.53),
             alpha = 0.1, color = "deeppink", 
             size = 10, pch = 16)+
  geom_text(aes(x = -0.0984, y = 51.5138, label = "St Paul’s"), 
            colour = "yellow", size = 5)
```

`alpha` is transparency (1 = opaque, 0 = invisible!)

We can also give R a dataframe with the latitude and longitude of multiple places

Let's plot the underground stations on our map:
```{r, message = FALSE, warning = FALSE}
underground <- read.csv("data-for-stats/underground.csv")
glimpse(underground)

ggmap(mapLondon, extent = "device") +
  geom_point(aes(x = Longitude, y = Latitude),
             data = underground, alpha = 0.8, color = "darkred")
```


We can scale the point sizes by a numeric variable...
```{r, message = FALSE, warning = FALSE}
ggmap(mapLondon, extent = "panel") + # We now use panel so we can see the legend
    geom_point(aes(x = Longitude, y = Latitude, size = NumberLines),
               data = underground, alpha = 0.8, color = "darkred") +
    labs(x = "Longitude", y = "Latitude") # adds label names
```

Or change the color gradient by a numeric variable...
```{r, message = FALSE, warning = FALSE}
ggmap(mapLondon, extent = "panel") +
  geom_point(aes(x = Longitude, y = Latitude, color = NumberLines),
             data = underground, alpha = 0.8, size = 5) +
  labs(x = "Longitude", y = "Latitude")
```

We can also annotate points (looks a bit messy with this dataset)
```{r, message = FALSE, warning = FALSE}
ggmap(mapLondon, extent = "panel") +
  # add the points, size differs by NumberLines
  geom_point(aes(x = Longitude, y = Latitude, size = NumberLines),
             data = underground, alpha = 0.8, color = "darkred") +
  # add the text, info from underground data frame
  geom_text(aes(x = Longitude, y = Latitude, label = Station),
            data = underground, colour = "lightblue", size = 2) +
  labs(x = 'Longitude', y = 'Latitude') + 
  ggtitle("Underground stations in East London") # adds a title
```


We can also join point using lines using `geom_path`. This joins two defined stations:
```{r, message = FALSE, warning = FALSE}
ggmap(mapLondon, extent = "device") + 
  geom_point(aes(x = Longitude, y = Latitude), 
             data = underground, alpha = 0.8, color = "darkred") +
  geom_path(aes(x = Longitude, y = Latitude), 
            data = filter(underground, Station == "Bethnal Green" | Station == "Embankment"),
            color = "cornflowerblue", size = 1)
```

Can also join everything but it looks hideous (or glorious!) in this case!
```{r, message = FALSE, warning = FALSE}
ggmap(mapLondon, 
	  extent = "device") +
geom_point(aes(x = Longitude, y = Latitude), 
	data = underground, alpha = 0.8, color = "yellow") +
geom_path(aes(x = Longitude, y = Latitude), data = underground, color = "hotpink", size = 1)
```

To get distances between places you can use Google via `mapdist`. Note that you have a 2500 query limit per day You can see how many you have left using:
```{r, eval = FALSE}
distQueryCheck()
```

To get distances between places:
```{r, message = FALSE, warning = FALSE}
mapdist("London, UK", "Zurich, Switzerland", mode = "driving")
mapdist("London, UK", "Sheffield, UK", mode = "walking")
mapdist("Sheffield, UK", "Zurich, Switzerland", mode = "bicycling")
```
