---
title: "Introduction to data structures and hygiene"
author: "Sarah Supp"
date: "5/16/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


**There is a presentation which goes with this handout, use it to add notes if you need to.**

Before we start, remember we need to do a couple of things...


#### 1. Tell R where your data are
See the introduction handout if you can't remember how.

Check it worked using:
```{r}
list.files()
```
or look at the Files tab

#### 2. Load the packages needed
If you downloaded the entire "tidyverse" then you should already have the ggplot2 package. If not, then you will need to get it using "install.packages"
```{r}
library(tidyverse)
library(ggplot2)
```

#### 3. Import the data
First we are going to work with the dataset `MOMv3.3.txt`, which contains the body mass of all mammals on Earth (MOM = Masses Of Mammals). You can find the original source by Smith et al. 2003, [here](http://esapubs.org/archive/ecol/E084/094/)
```{r}

```

#### 4. Look at the data
```{r}

```

### What do I need to know about my data? How can that help me code? Work a project?


