---
title: "Activity for data wrangling and tidying"
author: "Sarah Supp"
date: "5/16/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Let's practice our skills with a real dataset!

### Portal Project Dataset 2.0
Now that we've thought about some general data rules and pitfalls, let's look at another version of the Portal dataset, but let's pull it into R this time.

[Access the Portal Teaching Dataset](https://figshare.com/articles/Portal_Project_Teaching_Database/1314459)

#### 1. Tell R where your data are
```{r}
library(tidyverse)
surveys <- read_csv("example_data/surveys.csv")
```


#### 4. Look at the data
```{r}
glimpse(surveys)
```

### Is this data tidy? 
(10 minutes)

* what do the columns and rows represent?
* how do you feel about the way date is recorded? Why?
* does missing data make the data messy?
* what would be our first steps to cleaning up this data for analysis?

***
Build some fairly open-ended tasks to build on things we've learned for:

* data exploration (summary, glimpse, range, unique, etc.)
* data visualization (ggplot)
* data wrangling (dplyr)

