---
title: "R Notebook"
output: html_notebook
---

## Setup and libraries
```{r}
library(tidyverse)

#-------- load data
species <- read_csv("example_data/species.csv")
surveys <- read_csv("example_data/surveys.csv")

glimpse(species)
glimpse(surveys)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

## Let's explore and summarize the Portal Project data. 

1. How many species are represented?
2. What years was the site surveyed?
3. What is the range of weights? hindfoot lengths?
4. What is the average weight? hindfoot length?

```{r}
surveys %>%
  summarise(num_years = max(year) - min(year),
            start_year = min(year),
            end_year= max(year),
            num_species = n_distinct(species_id),
            min_wgt = min(weight, na.rm=TRUE),
            mean_wgt = mean(weight, na.rm=TRUE),
            max_wgt = max(weight, na.rm=TRUE),
            min_hfl = min(hindfoot_length, na.rm=TRUE),
            mean_hfl = mean(hindfoot_length, na.rm=TRUE),
            max_hfl = max(hindfoot_length, na.rm=TRUE))
```

## Let's visualize relationships in the data

1. Is weight correlated with hindfoot length?

```{r}
ggplot(surveys, aes(hindfoot_length, weight, group=species_id)) + 
  geom_point(alpha=0.25) + 
  stat_smooth(method=lm, aes(col=species_id))
```

2. Look at a distribution of the weights by each species separately.
```{r}
ggplot(surveys, aes(weight)) + 
  geom_histogram() + 
  facet_wrap(~species_id)
```

## Merge the two datasets to make it easier
```{r}
surveys2 <- merge(surveys, species, all.x=T)
```

## Wrangle the data

```{r}
# Keep only the rodent species
rats <- surveys2 %>%
  filter(taxa == "Rodent")

# Count the number of species captured at the site in each year
species_richness <- rats %>%
  group_by(year) %>%
  summarise(total=n_distinct(species_id))

# Summarize the number and average weights of each species in each year and plot
summary_by_year_plot <- rats %>%
  group_by(year, plot_id, species_id) %>%
  summarise(count=n(), mean=mean(weight, na.rm=TRUE))

# Keep only complete records (no missing values)
complete_records <- summary_by_year_plot %>%
  filter(!is.na(mean))

# Change the species codes from "PB" to "CB" and "PP" to "CP". 
# Remove records only identified only to genus ("X").
rats2 <- rats %>%
  mutate(species_id = replace(species_id, species_id=="PB", "CB")) %>%
  mutate(species_id = replace(species_id, species_id=="PP", "CP"))

# summarize by year
rats2 <- rats2 %>%
  group_by(year, species_id) %>%
  summarise(count=n())
```

## More data visualization on the cleaned and summary datasets

Which rodent species have been captured most frequently
```{r}
# Which rodent species have been captured most frequently?
ggplot(summary_by_year_plot, aes(year, count, group=species_id)) + 
  geom_line(aes(col=species_id))
```


### Has the number of species at the site changed through time?
```{r}
richness_thru_time <- ggplot(species_richness, aes(year, total)) + 
  geom_line(col="deeppink")
```

### Has average weight changed through time?
```{r}
ggplot(summary_by_year_plot, aes(year, mean, col=species_id)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~plot_id)
```

### Summarize the average weights of each sex and species
```{r}
average_MF_mass <- rats %>%
  filter(species_id %in% c("DM", "DO", "CB", "CP")) %>%
  group_by(sex, species_id) %>%
  summarise(mean_mass=mean(weight, na.rm=TRUE))  

ggplot(average_MF_mass, aes(sex, mean_mass)) + 
  geom_boxplot(aes(col=sex)) + facet_wrap(~species_id)
```

#![This is a cat](http://placekitten.com/600/300)

#![And this is a bear](http://placebear.com/600/300)





