---
title: "Activity for data wrangling and tidying"
author: "Sarah Supp"
date: "5/16/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Let's practice our skills with a real dataset!

### Portal Project Dataset 2.0
Now that we've thought about some general data rules and pitfalls, let's look at another version of the Portal dataset, but let's pull it into R this time.

[Access the Portal Teaching Dataset](https://figshare.com/articles/Portal_Project_Teaching_Database/1314459)

#### 1. Tell R where your data are
```{r}
library(conflicted)
library(tidyverse)
surveys <- read_csv("example_data/surveys.csv")
species <- read_csv("example_data/species.csv")
```

#### 2. Look at the data
```{r}
glimpse(surveys)
glimpse(species)
```

### Is this data tidy? 
(10 minutes)

* what do the columns and rows represent?
* how do you feel about the way date is recorded? Why?
* does missing data make the data messy?
* what would be our first steps to cleaning up this data for analysis?

***
Let's try some open-ended tasks to build on things we've learned so far:

### data exploration (summary, glimpse, range, unique, etc.)

1. How many species are represented? How many of these are rodents?
2. What years was the site surveyed?
3. What is the range of weights? hindfoot lengths?
4. What is the average weight? hindfoot length?

### data visualization 1.0 (ggplot)

1. Is weight correlated with hindfoot length?
2. Look at a distribution of the weights by each species separately.

### data wrangling (dplyr)

1. Keep only the rodent species.
2. Make a second table that counts the number of species captured at the site in each year. 
3. Make a third table that summarizes the number of individuals of each species captured in each year on each plot and their average weights.
4. Make a fourth table that only keeps complete records (no missing values)
5. The species "PB" stands for *Perognathus baileyi* which is now in the genus *Chaetodipus* (also true for "PP", *Perognathus penicillatus*. Change the species codes from "PB" to "CB" and "PP" to "CP". 
6. Remove records where the individual was identified only to genus.
7. In the actual dataset, *Neotoma albigula* is identified as NA. Why do you think it was changed to NL here? Could you change it back to NA? Why or why not?

### data visualization 2.0 (ggplot)

1. Which rodent species have been captured most frequently?
2. Has the number of species at the site changed through time?
3. Has average weight changed through time?
4. Are the average weights different for males of females of each species?

### What else would you like to look at in this dataset? Explore!
There are many different kinds of plots we can make, and different ways we can add third, fourth, or more variables to our plots using color, shape, size, facets, and more. 

1. Go to [Top 50 R plots](http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html) and scan through for a plot that looks interesting to you. Code is provided on this website. See if you can translate it to making a similar looking plot using this dataset.

2. Want a certain *style* to your plots that matches the journal specifications, or looks cool for a presentation? Try using ggthemes. Scan through the options on the [ggthemes website](https://cran.r-project.org/web/packages/ggthemes/vignettes/ggthemes.html) and try one or more that look appealing to you.

3. There are even more ways to polish your plots, by choosing [custom colors or palettes](https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/colorPaletteCheatsheet.pdf), using [color-blind friendly colors](https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html), or using additional packages.


